---
title: "Analyse Decathlon data using PCA"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  This tutorial analyses performance data during the 2004 Olympic Game.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(FactoMineR)
library(Rfast)
library(factoextra)
knitr::opts_chunk$set(echo = FALSE)
data("decathlon")
```

## Welcome

This tutorial will use data from athlete performance during the 2004 Olympic Game can be analysed to extract information that might not be obvious at first glance.

This tutorial uses a programming language called _R_ which was developed for statistical analyses.
It is widely used in a range of domains, from banking to biomedical data analyses.

In this tutorial, each screen will present you with some information, and code chunks for entering code to complete the exercises.

## Loading the data

### Exercise

The dataset that we will analyse is provided by a package called [FactoMineR](https://cran.r-project.org/package=FactoMineR),
in an object called `decathlon`.

Whatever dataset you work with, your first move is usually to have a look at it,
to get familiar with its contents.

In _R_, typing the name of an object displays the value of that object.

Since the object is called `decathlon`, type `decathlon` in the code chunk below,
then click the 'Run code' button on the right.

```{r decathlon, exercise=TRUE}

```

Take a moment to think about what you are seeing above:

- Can you describe in your own words what this dataset it showing? 
- What do the rows represent?
- What do the columns represent?
- What type of data is present in each column (is it numbers, text, other?)

When you're ready, move on to the next section where we describe the dataset in our own words.

## The decathlon dataset

This dataset contains the results of decathlon events during two athletic meetings which took place one month apart in 2004: the Olympic Games in Athens (23 and 24 August), and the Decastar 2004 (25 and 26 September).

For both competitions, the following information is available for each athlete:

- performance for each of the 10 events   
- total number of points (for each event, an athlete earns points based on performance; this is the sum of points scored) 
- final ranking

Each athlete is represented by a row in the table.

Each column represents one of the events, except for the last three columns which represent the overall rank, points, and the competition.

## Working with data frames in _R_

In _R_, a dataset shaped like a table (with rows and columns) is called a dataframe.

One of the most common things you will do when working with dataframes is accessing specific columns of data, either to analyse them, or the visualise them.

To access a specific column of the table, you type the name of the table, followed by the "dollar" sign `$` and the name of the column that you want to access.
For instance, you can get the values in the second column called `Long.jump` by typing `decathlon$Long.jump`.
Try it below!

```{r decathlon-long-jump, exercise=TRUE}

```

Notice how the code above only displays the values in that column,
without the name of the column or the names of the athletes those numbers correspond to.

Also, take note that things get a bit more complicated if the name of the column starts with a digit,
for instance the first column called `100m`.
In that case, you need to add "backticks" around the name of the column, like so:

```{r decathlon-100m, exercise=TRUE}
decathlon$`100m`
```

## Visualising data

Now that you know how to access values in specific columns,
you're only one step away from making graphs (also called "plots") that can help you visualise those values.

There are a number of simple plotting functions available in _R_ that can get you started quickly.
We'll stick with those in this tutorial to keep the code simple,
but there are more complicated functions that can produce much prettier versions of the same graphics.

For instance, the `plot()` function can be given a set of values between the parentheses,
and displays them in order, at a height that corresponds to each value.

You can try it by typing `plot(decathlon$Javeline)` in the code chunk below.

```{r plot-javelin, exercise=TRUE, fig.height=7, fig.width=9}

```

This is a bit hard to read,
let's use the `sort()` function to rearrange the values in increasing order before plotting them.

Here, we give you the code.
Can you describe how it differs from the code above?

```{r plot-sort-javelin, exercise=TRUE, fig.height=7, fig.width=9}
plot(sort(decathlon$Javeline))
```

Another plotting function is `hist()` which takes a set of values between the parentheses,
and displays a histogram that shows you how many values it counted in a series of "bins" of equal sizes
going from the smallest to the highest value observed.
You can try it by typing `hist(decathlon$Javeline)` in the code chunk below:

```{r hist-javelin, exercise=TRUE, fig.height=7, fig.width=9}

```

Take a moment to think about what those graphs represent,
what they are telling you about the data.

When you are ready, continue reading below.

- The first and second graphs show us the distance each competitor threw the javelin.
- The last graph shows the number of competitors who threw the javelin in groups of 5 metres.

## Summarising data

The _R_ programming language is full of functions that can analyse and summarise vast amounts of data.

For example, you might want to ask:

- What is the mean distance javelins were thrown?
- What is the furthest a javelin was thrown?

In the code chunks below, use the functions `mean()` and `max()` to answer those two questions.

Hint: the numbers you want to analyse are in the column named `Javeline` in the `decathlon` dataset.

```{r mean-javeline, exercise=TRUE}
# Use this code chunk to calculate the mean

```

```{r max-javeline, exercise=TRUE}
# Use this code chunk to calculate the maximum

```

### Annotating visualisations

Now that you know how to calculate the mean value for the javeline event,
you might want to highlight that value in the graph that we made previously to show all athletes.

To achieve this, you need to do two things:

- Copy the earlier code to make the plot showing each athlete, in sorted order.
- Add a horizontal line at the value of the mean, using the `abline()` function.

Note that the `abline()` function is very flexible, and can draw many different types of lines.
To clarify that we want a *horizontal* line,
we need to use the `h=` argument of the function and give it the value at which we want to draw that horizontal line, like so:

```{r plot-abline-javeline, exercise=TRUE}
plot(sort(decathlon$Javeline))
abline(h = mean(decathlon$Javeline))
```

Bonus points:

- Can you change the colour of the line?
- Can you add a title to the plot?

Edit the code chunk below to try!

```{r plot-abline-javeline-bonus, exercise=TRUE}
plot(sort(decathlon$Javeline))
abline(h = mean(decathlon$Javeline))
```

## Investigating hypotheses

Visualisation can be used to explore the data and investigate hypotheses you may have.

For instance, the `plot()` function can also be give two sets of values,
in which case it will draw a "scatterplot" that uses those two values as the coordinates to place each point.
For this to work, the two sets must have the same number of values.
Also, the two sets of values must be separated by a comma `,` within the parentheses of the function.

This is all a bit complicated to describe, so here we give you the code
and you only need to click the 'Run code' button.

```{r plot-100m-long-jump, exercise=TRUE, fig.height=7, fig.width=9}
plot(decathlon$`100m`, decathlon$Long.jump)
```

Take a moment to understand what the plot is showing.
What does each point represent?
Can you see any meaningful trend?
If so, how would you describe it?

- Each point represents an athlete.
- The position of each point along the horizontal axis shows the time (in seconds) it took them to run 100 meters.
- The position of each point along the vertical axis shows the distance the same athletes could jump in length (in meters).

It seems that the fastest an athlete can run, the furthest they can jump.
Of course, this is not surprising, but here we have data to prove it!

### Modelling data

In fact, we shouldn't trust our eyes alone, but rather formally analyse whether the trend is real.

One simple way to do that is to draw a line that best fits the data;
this is called a machine learning technique called "linear regression".

In _R_, this can be done using the `lm()` function.

The name of the function stands for "linear model".
It highlights that the technique "models" the data using a line which describes the relation between two variables.

To illustrate the technique, let's apply it to the same data as above.

The code is bit strange to type, so we're giving it to you here,
you only need to click 'Run code'.

To summarise:

- We use the `lm()` function to compute a linear model for the 100m and long-jump events.
- We use the `plot()` function to plot the performance of each individual athlete in those events.
- We add a (blue) line that represents the linear model computed in the first step.

```{r lm-100m-long-jump, exercise=TRUE, fig.height=7, fig.width=9}
linear_model <- lm(Long.jump ~ `100m`, data = decathlon)
plot(decathlon$`100m`, decathlon$Long.jump)
abline(linear_model, col = "blue")
```

What do you think about the plot and the trend represented by the linear model?

Although less exciting than a graph, it is possible to "see" the linear model from a numerical perspective.
Simply type the name of your linear model, to reveal its "coefficients".

```{r lm-100m-long-jump-print, exercise=TRUE}
linear_model <- lm(Long.jump ~ `100m`, data = decathlon)
linear_model
```

Take a moment to think about it.
A simple way to describe a line is using two coefficients:

- The "slope", which describes how quickly one value increases (or decreases) when the other one increase.
  In this case, how the distance an athlete can jump decreases, the longer it takes them to run 100 meters.
- The "intercept", which describes the "baseline" for one value when the other one is 0.
  In this case, the distance predicted by the model that an athlete would jump if it took them 0 seconds to run 100 meters.

Needless to say, in this case, the intercept value is nonsensical.
However, it is necessary to describe the linear model,
as there is a infinite number of lines with the same slope,
but only one that best fits the data.

### Measuring correlation

While the linear model above visually *suggests* that there is a significant relationship between
the time to run 100 meters and the distance athletes can jump,
it doesn't *prove* it.

Instead, we can numerically quantify the "correlation" between the two performances,
summarising the trend that we observed into a single number that measures how tightly
connected the two performances are.
This number is called the "correlation coefficient".

In _R_, the `cor()` function takes two sets of values and calculates
their correlation coefficient.

Try it below:

```{r cor-100m-long-jmup, exercise=TRUE}
cor(decathlon$`100m`, decathlon$Long.jump)
```

The value of a correlation coefficient can only be between `-1` and `1`.

- If it is exactly `1`,  the two sets of values are perfectly correlated.
  When one value increases by a certain amount, the other value always **increases** proportionally.
- If it is exactly `-1`, the two sets of values are perfectly "anti-correlated".
  When one value increases by a certain amount, the other one always **decreases** proportionally.
- If it is exactly `0`, the two sets of valyes are perfectly "uncorrelated".
  However one value changes, the other one changes in a random and unpredictable way.

What do you think about the value calculated above?
Can you describe how it relates to the graph further above?

## Preprocessing data

### Measuring variance

Variance is a mathematical measure of how spread out a set of number is.
It simple terms, it tells you how much the values in a dataset differ from the average (mean) value.

- If all the numbers are close to the mean, the variance is small (close to 0).
- If the numbers are more spread out, the variance is larger (the number can be infinitely large).

The variance is always positive, the minimum value of variance is exactly 0 if all the numbers are identical.

The `var()` function can be given a set of numbers, and calculates the variance for that set.
For instance, type `var(decathlon$Javeline)` in the code chunk below to calculate the variance
in the distance that atheletes have thrown javelins.

```{r var-decathlon-javeline, exercise=TRUE}

```

Alternatively, the `colVars()` function can be given a whole dataframe,
and will automatically calculate the variance in each column of the dataframe separately.
Try it below by typing `colVars(decathlon)`.

```{r colvars-decathlon, exercise=TRUE}

```

### Scaling data

Often, datasets contain data with different levels of variance, which makes them difficult to compare fairly.

For instance, in the previous exercise, we can see that
the "100m" event has a variance of 0.07
while the Javeline event has a variance of approximately 23.

To perform any analysis that compare the values of those two events fairly,
we should "scale" the data of the two events.

Briefly, "scaling" replaces a set of values into equivalent values that have a mean of 0 and a variance of 1.
It does it as follows:

- calculate the mean and the variance for a set of value.
- subtracts the mean from every value.
- divide every new value by the square root of the variance (a value called the "standard deviation").

Importantly, "scaling" does not change the *shape* of the data,
it only changes the numbers to be centered on 0 and spread within a similar distance from 0.

Let's illustrate it with an example.

In the first code chunk below, type `hist(decathlon$Javeline)` to display the distribution of distances at which javelins were thrown.

```{r hist-javelin-prescale, exercise=TRUE}

```

In the next code chunk, type `scaled_javeline <- scale(decathlon$Javeline)` to calculate scaled values for the Javelin performance.

```{r scale-javelin, exercise=TRUE}

```

Finally, in the next code chunk, type 

```{r hist-javelin-postscale, exercise=TRUE}

```

What differences do you see between the last two graphs?

- Most importantly, the values along the horizontal axis have change from [50, 75] to [-2, 3].
  It might not look like it is *visually* centered on 0,
  but that is because the original numbers are not distributed *symmetrically* around a mean value.
  You can trust the maths that the overall mean of the scaled values is indeed 0!
- You will also notice that the height of the bars have changed slightly,
  as if the *shape* of the data has changed.
  This is actually an artefact of the way the histogram creates "bins".
  Values near the edges of the bins can switch to a neighbouring bin when
  the values are scaled and the boundaries of the bins are calculated.

## Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

