---
title: "Analyse Decathlon data using PCA"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  This tutorial analyses performance data during the 2004 Olympic Game.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(FactoMineR)
library(Rfast)
library(factoextra)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome

This tutorial will use data from athlete performance during the 2004 Olympic Game can be analysed to extract information that might not be obvious at first glance.

This tutorial uses a programming language called _R_ which was developed for statistical analyses.
It is widely used in a range of domains, from banking to biomedical data analyses.

In this tutorial, each screen will present you with some information, and code chunks for entering code to complete the exercises.

## Loading the data

### Exercise

The dataset that we will analyse is provided by a package called [FactoMineR](https://cran.r-project.org/package=FactoMineR),
in an object called `decathlon`.

To load the dataset, type `data(decathlon)` in the code chunk below,
then click the 'Run code' button on the right.

```{r data-decathlon, exercise=TRUE}

```

At first glance, nothing seems to have happened.
However, as promised, the code above *loaded* the dataset in your _R_ session,
which means that it is now available for use.

Whatever dataset you work with, your first move is usually to have a look at it,
to get familiar with its contents.

In _R_, typing the name of an object displays the value of that object.

Since the object is called `decathlon`, type `decathlon` in the code chunk below,
then click the 'Run code' button on the right.

```{r decathlon, exercise=TRUE}

```

Take a moment to think about what you are seeing above:

- Can you describe in your own words what this dataset it showing? 
- What do the rows represent?
- What do the columns represent?
- What type of data is present in each column (is it numbers, text, other?)

When you're ready, move on to the next section where we describe the dataset in our own words.

## The decathlon dataset

This dataset contains the results of decathlon events during two athletic meetings which took place one month apart in 2004: the Olympic Games in Athens (23 and 24 August), and the Decastar 2004 (25 and 26 September).

For both competitions, the following information is available for each athlete:

- performance for each of the 10 events   
- total number of points (for each event, an athlete earns points based on performance; this is the sum of points scored) 
- final ranking

Each athlete is represented by a row in the table.

Each column represents one of the events, except for the last three columns which represent the overall rank, points, and the competition.

## Working with data frames in _R_

In _R_, a dataset shaped like a table (with rows and columns) is called a dataframe.

One of the most common things you will do when working with dataframes is accessing specific columns of data, either to analyse them, or the visualise them.

To access a specific column of the table, you type the name of the table, followed by the "dollar" sign `$` and the name of the column that you want to access.
For instance, you can get the values in the second column called `Long.jump` by typing `decathlon$Long.jump`.
Try it below!

```{r decathlon-long-jump, exercise=TRUE}

```

Note that things get a bit more complicated if the name of the column starts with a digit, for instance the first column called `100m`.
In that case, you need to "backticks" around the name of the column, like so:

```{r decathlon-100m, exercise=TRUE}
decathlon$`100m`
```

Thankfully, the RStudio program that we are using to run this tutorial makes coding easier.
As soon as you type the name of a dataframe followed by the dollar sign,
it opens a dropdown menu with the list of column names available,
and autocompletes the correct code for whatever column name you select.

Try it below for the column named `400m`!

```{r decathlon-400m, exercise=TRUE}

```

## Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

