---
title: "Analyse Decathlon data using PCA"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  This tutorial analyses performance data during the 2004 Olympic Game.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(FactoMineR)
library(Rfast)
library(factoextra)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome

This tutorial will use data from athlete performance during the 2004 Olympic Game can be analysed to extract information that might not be obvious at first glance.

This tutorial uses a programming language called _R_ which was developed for statistical analyses.
It is widely used in a range of domains, from banking to biomedical data analyses.

In this tutorial, each screen will present you with some information, and code chunks for entering code to complete the exercises.

## Loading the data

### Exercise

Actually, the dataset that we will analyse is already loaded.
That dataset is provided by a package called [FactoMineR](https://cran.r-project.org/package=FactoMineR),
in an object called `decathlon`.

In _R_, typing the name of an object displays the value of that object.

To demonstrate, the code chunk below is already filled with the code `decathlon`.
You only need to click the button 'Run code'.

```{r decathlon, exercise=TRUE}
decathlon
```

You can also be more explicit using the `print()` function which displays the value of any object given to it using the parentheses.

Again, the code is given to you here.
You only need to click the button 'Run code'.

```{r print-decathlon, exercise=TRUE}
print(decathlon)
```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

